Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)

### 2014-02-07 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 3 Keynote:  Minor changes.

The first
keynote~\cite{WSSSPE1-keynote1}
was delivered by Philip E. Bourne of University of California, San Diego.  Bourne is a basic biomedical scientist but has formed four software
companies. He also helped found PLOS and
PDB~\cite{pdb-web}.  He is
working on automating three-dimensional visualizations of cell
contents and molecular structures, a problem that has not been solved
and would serve as a key function of software in biomedical sciences.

Bourne's keynote presentation is entiled ``A Recipe for Sustainable Software''.
He emphasized that sustainability for software ``does not just
mean more money from Government.''  He noted that in places,
development in science has improved thanks to open source and Github,
but for the most part remains arcane. He argued that there is much we
can learn from the App Store model. He mentioned
BioJava~\cite{biojava-web} and
OpenScienceDataCloud~\cite{osdc-web}
as distribution channels.  On a related note, Bourne observed a common
evolutionary pathway for computational biology projects, from data
archive to analytics platform to educational use, and suggested that
use of scientific software for outreach might be the final step.

@@

Another problem he experienced was staff retention with respect to
private organizations which reward those combining research and
software expertise (the ``Google Bus''). However, he is a strong fan
of private sector sustainability sources and public-private
partnerships. He noted that making a successful business from
scientific software alone is very rare: founders overvalue while
customers undervalue. He noted that to last, an open source project
needs a minimal funding requirement even with a vibrant community --
good will only goes so far if you are being paid to do something else.
He talked about grant schemes of relevance in the U.S., particularly
with regard to technology transfer.~\cite{sbir-web, fased-web}.

@@

His talk used an example from data reduction in Astronomy, where he
needed to remove interfering effects from the device. He built a ``bad
pixel mask'', and realized that while this is persistent, there was no
way/practice of sharing this with the data, so many researchers
repeated the same calculations. He estimated that 13 person-years were
wasted in this redundant calculation.

@@

C. Titus Brown~\cite{ged-web} argues that with
regard to reusable software, ``we should just start doing it.'' Arfon
noted that documentation should be ``treated as a first class
entity.''  He noted that some software fields have excellent cultures
of module reuse, such as
RubyGems~\cite{rubygems-web},
PyPI~\cite{pypi-web}, and
CPAN~\cite{cpan-web}, where there is effectively
ultra-low-friction collaboration through these repositories. This has
not happened in highly numerical, compiled language scientific
software.

@@

Fernando Perez says that ``Open Source is reproducible by necessity''
- since reproducibility is a prerequisite for remote collaboration.
Arfon therefore defined GitHub as part of the next stage in
development of the web, following on from the ``social web'' of
Facebook et.al.: ``The Collaborative Web.''

>>>>

The first
keynote~\cite{WSSSPE1-keynote1}
was delivered by Philip E. Bourne of University of California, San Diego.  Bourne is a basic biomedical scientist but has formed four software
companies. He also helped found PLOS and
PDB~\cite{pdb-web}.  He is
working on automating three-dimensional visualizations of cell
contents and molecular structures, a problem that has not been solved
and when done, would serve as a key function of software in biomedical sciences.

Bourne's keynote presentation is entitled ``A Recipe for Sustainable Software''.
He emphasized that sustainability for software ``does not just
mean more money from Government.''  He noted that in places,
development in science has improved thanks to open source and Github,
but for the most part remains arcane. He argued that there is much we
can learn from the App Store model. He mentioned
BioJava~\cite{biojava-web} and
OpenScienceDataCloud~\cite{osdc-web}
as distribution channels.  On a related note, Bourne observed a common
evolutionary pathway for computational biology projects, from data
archive to analytics platform to educational use, and suggested that
use of scientific software for outreach might be the final step.

@@

Another problem he experienced was staff retention with respect to
private organizations which reward those combining research and
software expertise (the ``Google Bus''). However, he is a strong fan
of private sector sustainability sources and public-private
partnerships. He noted that making a successful business from
scientific software alone is very rare: founders overvalue while
customers undervalue. He noted that to last, an open source project
needs a minimal funding requirement even with a vibrant community --
good will only goes so far if you are being paid to do something else.
He talked about grant schemes of relevance in the U.S., particularly
with regard to technology transfer~\cite{sbir-web, fased-web}.

@@

His talk used an example from data reduction in Astronomy, where he
needed to remove interfering effects from the device. He built a ``bad
pixel mask'', and realized that while it was persistent, there was no
way/practice of sharing this with the data, so many researchers
repeated the same calculations. He estimated that 13 person-years were
wasted in this redundant calculation.

@@

C. Titus Brown~\cite{ged-web} argued that with
regard to reusable software, ``we should just start doing it.'' Smith
noted that documentation should be ``treated as a first class
entity.''  He noted that some software fields have excellent cultures
of module reuse, such as
RubyGems~\cite{rubygems-web},
PyPI~\cite{pypi-web}, and
CPAN~\cite{cpan-web}, where there is effectively
ultra-low-friction collaboration through these repositories. This has
not happened in highly numerical, compiled language scientific
software.

@@

Fernando Perez emphasize that ``open source is reproducible by necessity''
- since reproducibility is a prerequisite for remote collaboration.
Smith therefore defined GitHub as part of the next stage in web
development, following on from the social web of
Facebook et al.: ``The Collaborative Web.''
